quollr: An R Package for Visualizing 2-D Models from Non-linear Dimension Reductions in High Dimensional Space

### 2025-09-18 paper-quollr.Rmd
Updated notation table

```{r overview, eval=knitr::is_latex_output(), fig.cap="Wireframe model representation of the NLDR layout, lifted and displayed in high-dimensional space. The left panel shows the NLDR layout with a triangular mesh overlay, forming the wireframe structure. This mesh can be lifted into higher dimensions and projected to examine how the geometric structure of the data is preserved. Panels (a1–a4) display different $2\\text{-}D$ projections of the lifted wireframe, where the underlying curved sheet structure of the data is more clearly visible. The triangulated mesh highlights how local neighborhoods in the layout correspond to relationships in the high-dimensional space, enabling diagnostics of distortion and preservation across dimensions.", fig.pos='H'}
hex_grid_scurve + wrap_plots(
  scurve_proj_umap_model1, scurve_proj_umap_model2,
  scurve_proj_umap_model3, scurve_proj_umap_model4, 
  ncol = 2)
```

@@

```{r, echo=TRUE, eval=FALSE}
fit_highd_model(
  highd_data = scurve, 
  nldr_data = scurve_umap, 
  b1 = 25, 
  q = 0.1, 
  hd_thresh = 1)
```

@@

Constructing the $2\text{-}D$ model primarily involves (i) scaling the NLDR data, (ii) binning the data, (iii) obtaining bin centroids, (iv) connecting centroids with line segments to indicate neighbors, and (v) removing low-density hexagons.

@@

notation_df |> 
  kable(caption = "Summary of notation for describing new methodology.", booktabs = T) |>
  column_spec(1, width = "3cm") |>
  column_spec(2, width = "12cm")
```

>>>>

```{r overview, eval=knitr::is_latex_output(), fig.cap="Wireframe model representation of the NLDR layout, lifted and displayed in high-dimensional space. The left panel shows the NLDR layout with a triangular mesh overlay, forming the wireframe structure. This mesh can be lifted into higher dimensions and projected to examine how the geometric structure of the data is preserved. Panels (a1–a4) display different \\twoD projections of the lifted wireframe, where the underlying curved sheet structure of the data is more clearly visible. The triangulated mesh highlights how local neighborhoods in the layout correspond to relationships in the high-dimensional space, enabling diagnostics of distortion and preservation across dimensions.", fig.pos='H'}
hex_grid_scurve + wrap_plots(
  scurve_proj_umap_model1, scurve_proj_umap_model2,
  scurve_proj_umap_model3, scurve_proj_umap_model4, 
  ncol = 2)
```

@@

```{r, echo=TRUE, eval=FALSE}
fit_highd_model(
  highd_data = scurve, 
  nldr_data = scurve_umap, 
  b1 = 21, 
  q = 0.1, 
  hd_thresh = 0)
```

@@

Constructing the \twoD model primarily involves (i) scaling the NLDR data, (ii) binning the data, (iii) obtaining bin centroids, (iv) connecting centroids with line segments to indicate neighbors, and (v) removing low-density hexagons.

@@

notation_df |> 
  kable(caption = "Summary of notation for describing new methodology.", booktabs = T) |>
  column_spec(1, width = "3cm") |>
  column_spec(2, width = "10cm") |>
  column_spec(3, width = "3cm")
```

### 2025-09-18 paper-quollr.tex
Updated notation table

\begin{figure}[H]
\includegraphics[width=1\linewidth]{paper-quollr_files/figure-latex/overview-1} \caption{Wireframe model representation of the NLDR layout, lifted and displayed in high-dimensional space. The left panel shows the NLDR layout with a triangular mesh overlay, forming the wireframe structure. This mesh can be lifted into higher dimensions and projected to examine how the geometric structure of the data is preserved. Panels (a1–a4) display different $2\text{-}D$ projections of the lifted wireframe, where the underlying curved sheet structure of the data is more clearly visible. The triangulated mesh highlights how local neighborhoods in the layout correspond to relationships in the high-dimensional space, enabling diagnostics of distortion and preservation across dimensions.}\label{fig:overview}
\end{figure}

@@

\begin{verbatim}
fit_highd_model(
  highd_data = scurve, 
  nldr_data = scurve_umap, 
  b1 = 25, 
  q = 0.1, 
  hd_thresh = 1)
\end{verbatim}

@@

Constructing the \(2\text{-}D\) model primarily involves (i) scaling the NLDR data, (ii) binning the data, (iii) obtaining bin centroids, (iv) connecting centroids with line segments to indicate neighbors, and (v) removing low-density hexagons.

@@

\caption{\label{tab:notation-tb-pdf}Summary of notation for describing new methodology.}
\centering
\begin{tabular}[t]{>{\raggedright\arraybackslash}p{3cm}>{\raggedright\arraybackslash}p{12cm}}
\toprule
Notation & Description\\
\midrule
\textbackslash{}\$n, p, k\textbackslash{}\$ & number of observations, variables, embedding dimension, respectively\\
\$\textbackslash{}mathbfit\{X\}, \textbackslash{}mathbfit\{x\}\$ & \$p\$-dimensional data (population, sample)\\
\$\textbackslash{}mathbfit\{y\}\$ & \$k\$-dimensional layout\\
\$P\$ & orthonormal basis, generating a \$d\textbackslash{}text\{-\}dimensional\$ linear projection of \$p\$-dimensional data\\
\$T\$ & true  model\\
\addlinespace
\$g\$ & functional mapping from \textbackslash{}pD\{\} to \textbackslash{}kD\{\}, especially as prescribed by NLDR\\
\$\textbackslash{}mathbfit\{\textbackslash{}theta\}\$ & (Hyper-) parameters for NLDR method\\
\$r\$ & ranges of the embedding components\\
\$C\textasciicircum{}\{(j)\}\$ & \$j\$-dimensional bin centers\\
\$(b\_1, b\_2)\$ & number of bins in each direction\\
\addlinespace
\$(a\_1, a\_2)\$ & binwidths, distance between centroids in each direction\\
\$(s\_1, \textbackslash{} s\_2)\$ & starting coordinates of the hexagonal grid\\
\$q\$ & buffer to ensure hexgrid covers data, proportion of data range, 0-1\\
\$m\$ & number of non-empty bins\\
\$b\$ & number of  hexagons in the grid\\
\addlinespace
\$h\$ & hexagonal id\\
\$l\$ & side length\\
\$A\$ & area\\
\$n\_h\$ & bin count\\
\$w\_h\$ & standardised bin counts\\
\bottomrule
\end{tabular}
\end{table}

>>>>

\begin{figure}[H]
\includegraphics[width=1\linewidth]{paper-quollr_files/figure-latex/overview-1} \caption{Wireframe model representation of the NLDR layout, lifted and displayed in high-dimensional space. The left panel shows the NLDR layout with a triangular mesh overlay, forming the wireframe structure. This mesh can be lifted into higher dimensions and projected to examine how the geometric structure of the data is preserved. Panels (a1–a4) display different $2\text{-}D$projections of the lifted wireframe, where the underlying curved sheet structure of the data is more clearly visible. The triangulated mesh highlights how local neighborhoods in the layout correspond to relationships in the high-dimensional space, enabling diagnostics of distortion and preservation across dimensions.}\label{fig:overview}
\end{figure}

@@

\begin{verbatim}
fit_highd_model(
  highd_data = scurve, 
  nldr_data = scurve_umap, 
  b1 = 21, 
  q = 0.1, 
  hd_thresh = 0)
\end{verbatim}

@@

Constructing the $2\text{-}D$ model primarily involves (i) scaling the NLDR data, (ii) binning the data, (iii) obtaining bin centroids, (iv) connecting centroids with line segments to indicate neighbors, and (v) removing low-density hexagons.

@@

\caption{\label{tab:notation-tb-pdf}Summary of notation for describing new methodology.}
\centering
\begin{tabular}[t]{>{\raggedright\arraybackslash}p{3cm}>{\raggedright\arraybackslash}p{10cm}>{\raggedright\arraybackslash}p{3cm}}
\toprule
Notation & Description & Variable\\
\midrule
\textbackslash{}\$n, p, k\textbackslash{}\$ & number of observations, variables, embedding dimension, respectively & NA\\
\$\textbackslash{}mathbfit\{X\}, \textbackslash{}mathbfit\{x\}\$ & \$p\$-dimensional data (population, sample) & NA\\
\$\textbackslash{}mathbfit\{y\}\$ & \$k\$-dimensional layout & NA\\
\$P\$ & orthonormal basis, generating a \$d\textbackslash{}text\{-\}dimensional\$ linear projection of \$p\$-dimensional data & NA\\
\$T\$ & true  model & NA\\
\addlinespace
\$g\$ & functional mapping from \textbackslash{}pD\{\} to \textbackslash{}kD\{\}, especially as prescribed by NLDR & NA\\
\$\textbackslash{}mathbfit\{\textbackslash{}theta\}\$ & (Hyper-) parameters for NLDR method & NA\\
\$r\$ & ranges of the embedding components & NA\\
\$C\textasciicircum{}\{(j)\}\$ & \$j\$-dimensional bin centers & NA\\
\$(b\_1, b\_2)\$ & number of bins in each direction & b1, b2\\
\addlinespace
\$(a\_1, a\_2)\$ & binwidths, distance between centroids in each direction & NA\\
\$(s\_1, \textbackslash{} s\_2)\$ & starting coordinates of the hexagonal grid & NA\\
\$q\$ & buffer to ensure hexgrid covers data, proportion of data range, 0-1 & q\\
\$m\$ & number of non-empty bins & NA\\
\$b\$ & number of  hexagons in the grid & NA\\
\addlinespace
\$h\$ & hexagonal id & NA\\
\$l\$ & side length & NA\\
\$A\$ & area & NA\\
\$n\_h\$ & bin count & NA\\
\$w\_h\$ & standardised bin counts & NA\\
\addlinespace
NA & threshold to extract high density hexagons & hd\_thresh\\
\bottomrule
\end{tabular}
\end{table}
